Dependent Object Types

### 2012-08-08 article.tex
reorg
proof

% Logical relations
\newcommand{\relv}[4]{\mathcal{V}_{#1;#2;#3}\llbracket#4\rrbracket}
\newcommand{\rele}[4]{\mathcal{E}_{#1;#2;#3}\llbracket#4\rrbracket}
\newcommand{\rels}[3]{\mathcal{S}_{#1;#2}\llbracket#3\rrbracket}
\newcommand{\relg}[3]{\mathcal{G}_{#1;#2}\llbracket#3\rrbracket}
\newcommand{\irred}[2]{\text{irred }(#1,#2)}
\newcommand{\andl}{\;\wedge\;}
\newcommand{\orl}{\vee}
\newcommand{\impliesl}{\rightarrow}
\newcommand{\reductionl}[5]{#1 \operatorname{|} #2 \;\rightarrow^{#5}\; #3 \operatorname{|} #4}
\newcommand{\ds}{\,\vDash\,}

@@

\subsubsection{$\mathcal{V}$}
$\relv k \Gamma s T$ defines the set of values that appear to have
type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
$\dom(\Gamma) = \dom(s)$ and $\forall (x : T) \in \Gamma, x \in \relv
k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s T$ simply
by being in the store. In addition, it belongs to $\relv k \Gamma s T$
for $k > 0$, if it defines all method and value labels in the
expansion of $T$ appropriately for $j < k$ steps.
\begin{align*}
&\relv k \Gamma s T = \{ y \ |\\
&\gap \Gamma \ts T \wfe \andl\\
&\gap y \in \dom(s) \andl\\
&\gap \forall j < k,\\
&\gap y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s,\\
&\gap \Gamma \ts T \expand_y \seq{D},\\
&\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
&\gap\gap t_i \in \rele j {\Gamma \envplus{x : S}} s U)\\
&\gap \andl\\
&\gap (\forall l_i : V \in \seq{D},\\
&\gap\gap v_i \in \relv j \Gamma s V)\\
&\}
\end{align*}

\subsubsection{$\mathcal{E}$}
$\rele k \Gamma s T$ defines the set of terms that appear to have type
$T$ when taking at most $k$ steps. $s$ must agree with $\Gamma$, but
$\Gamma$ can additionaly contain variables not in $s$. This is needed
for checking methods in $\mathcal{V}$ above, and for relating open
terms. If $k > 0$, $\mathcal{E}$ completes the store to agree with
$\Gamma$. It then states that if it can reduce $t$ in the completed
store to an irreducible term in $j < k$ steps, then this term must be
in a corresponding $\mathcal{V}$ set.
\begin{align*}
&\rele k \Gamma s T = \{ t \ |\\
&\gap \Gamma \ts T \wfe \andl\\
&\gap k = 0 \orl (\forall j \leq k,\\
&\gap\gap \forall s' \in \rels k \Gamma s ,\\
&\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
&\gap\gap \irred {t'} {s''} \impliesl\\
&\gap\gap \forall \Gamma' \in \relg k {s''} \Gamma ,\\
&\gap\gap {t'} \in \relv {k-j-1} {\Gamma'} {s''} T)\\
&\}
\end{align*}

\subsubsection{$\mathcal{S}$}
$\rels k \Gamma s$ for $k > 0$ defines the set of completed stores
that agree with $\Gamma$ on each variable when taking at most $k-1$
steps.
\begin{align*}
&\rels k \Gamma s = \{ s, \seq{x_i \mapsto c_i}^{m \leq i < n} \ |\\
&\gap s = \seq{x \mapsto c}^m \andl \Gamma = \seq{x : T}^n \andl \\
&\gap m\leq n \andl \forall i, m \leq i < n,\\
&\gap c_i \in \relv {k-1} \Gamma {s, \seq{x_j \mapsto c_j}^{m \leq j < i}} {T_i}\\
&\}
\end{align*}

\subsubsection{$\mathcal{G}$}
$\relg k s \Gamma$ defines a singleton set of a completed environment
that agrees with a store $s$ by simply copying the constructor type
from the store for each missing variable.
\begin{align*}
&\relg k s \Gamma = \{ \Gamma, \seq{x_i : T_{c_i}}^{m \leq i < n} \ |\\
&\gap \Gamma = \seq{x : T}^m \andl s = \seq{x \mapsto c}^n \andl\\
&\}
\end{align*}

@@

\subsection{Proof of the Fundamental Theorem}

@@

\begin{myproof}
The proof is on induction on the derivation of $\Gamma \ts t : T$. For
each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall k$. The
non-trivial case is to need to show ${t'} \in \relv {k-j-1} {\Gamma'}
{s'} T$ by supposing $k > 0$, $s \in \rels k \Gamma \emptyset$, and, for some
$j \leq k$, $\reductionl {t} s {t'} {s'} j \andl \irred {t'}
{s'} \andl \Gamma' \in \relg \Gamma k {s'} \Gamma$.

@@

\Case \textsc{var}: $\Gamma \ts x : T$ knowing $(x : T) \in
\Gamma$. $x \in \relv {k-1} \Gamma s T$ follows from the definition of
$\mathcal{G}$.

@@

%Need to show $t_1.l_i \in \rele k \Gamma \emptyset T$. Suppose $k > 0$ and $s
%\in \rels k \Gamma \emptyset$. Suppose, for some $j \leq k$, $\reductionl
%    {t_1.l_i} s {t'} {s'} j \andl \irred {t'} {s'} \andl \Gamma' \in
%    \relg \Gamma k {s'} \Gamma$. Remains to show ${t'} \in \relv
%          {k-j-1} {\Gamma'} {s'} T$.

@@

By operational semantics and the above, $\reductionl {t_1'.l_i} {s'}
{t'} {s'} 1$, and we can conclude $t' \in \relv {k-j-1} {\Gamma'} {s'}
T$ from the clause for value labels of ${t_1'} \in \relv {k-j}
{\Gamma'} {s'} {T_1}$.

@@

%Need to show $t_1.m_i(t_2) \in \rele k \Gamma \emptyset T$. Suppose $k > 0$
%and $s \in \rels k \Gamma \emptyset$. Suppose, for some $j \leq k$,
%$\reductionl {t_1.m_i(t_2)} s {t'} {s'} j \andl \irred {t'} {s'} \andl
%\Gamma' \in \relg \Gamma k {s'} \Gamma$. Remains to show ${t'} \in
%\relv {k-j-1} {\Gamma'} {s'} T$.

@@

Because $t_2$ reduces to a value $t_2'$ starting in store $s$, it
should also reduce to a value $v_2$ in the same number of steps
starting in store $s_1$, since $s_1$ includes $s$. So let $\reductionl
{t_2} {s_1} {v_2} {s_{12}} {j_2}$ with $v_2 \in \relv {k-j_2-1}
{\Gamma_{12}} {s_{12}} {T_2}$.

@@

%Need to show $\new y c {t'} \in \rele k \Gamma \emptyset T$. Suppose $k > 0$
%and $s \in \rels k \Gamma \emptyset$. Suppose, for some $j \leq k$,
%$\reductionl {\new y c {t'}} s {t''} {s''} j \andl \irred {t''} {s''} \andl
%\Gamma'' \in \relg \Gamma k {s''} \Gamma$. Remains to show ${t''} \in
%\relv {k-j-1} {\Gamma''} {s''} T$.

@@

\subsection{Proof of the Substitution Lemma}

The substitution lemma states that if (1) $v_2 \in \relv {k_2}
{\Gamma_{12}} {s_{12}} {T_2}$ and (2) $t \in \rele {k_1} {\Gamma_1
  \envplus{{x_i} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
(4) $x_i \not\in \fn(T)$ and $\Gamma$ included in $\Gamma_1$ and
$\Gamma_{12}$ included in $\Gamma_1$ and $s_{12}$ included in $s_1$,
then $\subst {v_2} {x_i} {t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}}
{s_{12}} T$.

\begin{proofsketch}
By (1) and (3), it should hold that (5) $v_2 \in \relv {k_2}
{\Gamma_{12}} {s_{12}} {S}$.  Since (2) holds, it should also hold
that ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x_i} :
    S}} {s_{12}} T$ by monotonicity of the logical relation on all
subscripts. Then, we can instantiate $x_i$ in the complete store to
map to what $v_2$ maps to. This should be fine by (5) and
monotonicity. Thus, ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
  \envplus{{x_i} : S}} {s_{12} \envplus{{x_i} \mapsto s_{12}(v)}}
T$. Thanks to (4), we don't actually need $x_i$ to be held abstract in
the environment, because it won't occur in $T$ or its expansion, so
we can use the actual type constructor of $v_2$, say $T_{v_2}$, instead
of $S$ for $x_i$: ${t_i} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
  \envplus{{x_i} : {T_{v_2}}}} {s_{12} \envplus{{x_i} \mapsto
      s_{12}(v)}} T$. This implies what needs to be shown.

@@

>>>>

% Logical relations
\newcommand{\relv}[4]{\mathcal{V}_{#1;#2;#3}\llbracket#4\rrbracket}
\newcommand{\rele}[4]{\mathcal{E}_{#1;#2;#3}\llbracket#4\rrbracket}
\newcommand{\rels}[3]{\mathcal{\supseteq}_{#1}\llbracket#2;#3\rrbracket}
\newcommand{\relg}[3]{\mathcal{\supseteq}_{#1;#2}\llbracket!#3\rrbracket}
\newcommand{\irred}[2]{\text{irred }(#1,#2)}
\newcommand{\andl}{\;\wedge\;}
\newcommand{\orl}{\vee}
\newcommand{\impliesl}{\rightarrow}
\newcommand{\reductionl}[5]{#1 \operatorname{|} #2 \;\rightarrow^{#5}\; #3 \operatorname{|} #4}
\newcommand{\ds}{\,\vDash\,}

@@

\subsubsection{Set of Values}
$\relv k \Gamma s T$ defines the set of values that appear to have
type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
$\dom(\Gamma) = \dom(s)$ (ordered) and $\forall (x : T) \in \Gamma, x
\in \relv k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s
T$ simply by being in the store. In addition, it belongs to $\relv k
\Gamma s T$ for $k > 0$, if it defines all method and value labels in
the expansion of $T$ appropriately for $j < k$ steps.
\begin{align*}
&\relv k \Gamma s T = \{ y \ |\\
&\gap \Gamma \ts T \wfe \andl\\
&\gap y \in \dom(s) \andl\\
&\gap \forall j < k,\\
&\gap y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s,\\
&\gap \Gamma \ts T \expand_y \seq{D},\\
&\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
&\gap\gap t_i \in \rele j {\Gamma \envplus{x : S}} s U)\\
&\gap \andl\\
&\gap (\forall l_i : V \in \seq{D},\\
&\gap\gap v_i \in \relv j \Gamma s V)\\
&\}
\end{align*}

\subsubsection{Set of Terms}
$\rele k \Gamma s T$ defines the set of terms that appear to have type
$T$ when taking at most $k$ steps. $s$ must agree with a {\it prefix}
of $\Gamma$, so $\Gamma$ can additionaly contain variables not in
$s$. This is needed for checking methods in $\mathcal{V}$ above, and
for relating open terms. If $k > 0$, $\mathcal{E}$ extends $\Gamma$
and $s$ so that they agree. It then states that if it can reduce $t$
in the extended store to an irreducible term in $j < k$ steps, then
this term must be in a corresponding $\mathcal{V}$ set.
\begin{align*}
&\rele k \Gamma s T = \{ t \ |\\
&\gap \Gamma \ts T \wfe \andl\\
&\gap k = 0 \orl (\forall j \leq k,\\
&\gap\gap \forall (\Gamma'; s') \in \rels k \Gamma s ,\\
&\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
&\gap\gap \irred {t'} {s''} \impliesl\\
&\gap\gap \forall \Gamma'' \in \relg k {s''} {\Gamma'} ,\\
&\gap\gap {t'} \in \relv {k-j-1} {\Gamma''} {s''} T)\\
&\}
\end{align*}

\subsubsection{Extending the environment and the store}
$\rels k \Gamma s$ for $k > 0$ defines the set of completed
environment and stores that agree on $k-1$ steps, and that extend
$\Gamma$ and $s$. $s$ must agree with a {\it prefix} of $\Gamma$.
Both $\Gamma$ and $s$ are ordered maps. For $s$, $s'$ extends $s$ if
$s$ is a prefix of $s'$. For $\Gamma$, $\Gamma'$ extends $\Gamma$ if
we get back $\Gamma$ by keeping only the elements of $\Gamma'$ that
belong to $\Gamma$. Furthermore, a prefix of $\Gamma'$ agrees with
$s$.
\begin{align*}
&\rels k \Gamma s = \{\\
&\ (\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i < n; 0 \leq j < i_n}; s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n; 0 \leq j < i_n}) |\\
&\ s = \seq{x \mapsto c}^m \andl \Gamma = \seq{x : T}^n \andl \\
&\ m\leq n \andl \forall i, m \leq i < n, \forall {i_n}, j, 0 \leq j < i_n,\\
&\ \forall T_{ij}, c_{ij}, T_{i{(i_n-1)}} = T_i, \forall n' \leq n, i_n' \leq i_n,\\
&\ c_{ij} \in \relv {k-1} {\seq{x : T}^m, \seq{x_{ij} : T_{ij}}^{m \leq i' < n; 0 \leq j < i_n'}} {s, \seq{x_{ij} \mapsto c_{ij}}^{m \leq i < n'; 0 \leq j < i_n'}} {T_{ij}}\\
&\}
\end{align*}

\subsubsection{Completing the environment to agree with the store}
$\relg k s \Gamma$ defines a singleton set of a completed environment
that agrees with a store $s$ by simply copying the constructor type
from the store for each missing variable.
\begin{align*}
&\relg k s \Gamma = \{ \Gamma, \seq{x_i : T_{c_i}}^{m \leq i < n} \ |\\
&\gap \Gamma = \seq{x : T}^m \andl s = \seq{x \mapsto c}^n \andl\\
&\}
\end{align*}

@@

\subsection{Statements and Proofs}

\subsubsection{Fundamental Theorem}

@@

\begin{myproof}
The proof is on induction on the derivation of $\Gamma \ts t : T$. For
each case, we need to show $t \in \rele k \Gamma \emptyset T, \forall k$. The
non-trivial case is to need to show ${t'} \in \relv {k-j-1} {\Gamma''}
{s'} T$ by supposing $k > 0$, $(\Gamma'; s) \in \rels k \Gamma \emptyset$, and, for some
$j \leq k$, $\reductionl {t} s {t'} {s'} j \andl \irred {t'}
{s'} \andl {\Gamma''} \in \relg \Gamma k {s'} {\Gamma'}$.

@@

\Case \textsc{var}: $\Gamma \ts x : T$ knowing $(x : T) \in
\Gamma$. $x \in \relv {k-1} {\Gamma'} s T$ follows from the definition of
$\rels k \Gamma \emptyset$.

@@

@@

By operational semantics and the above, $\reductionl {t_1'.l_i} {s'}
{t'} {s'} 1$, and we can conclude $t' \in \relv {k-j-1} {\Gamma''} {s'}
T$ from the clause for value labels of ${t_1'} \in \relv {k-j}
{\Gamma''} {s'} {T_1}$.

@@

@@

Because $t_2$ reduces to a value $t_2'$ starting in store $s$, it
should also reduce to a value $v_2$ in the same number of steps
starting in store $s_1$, since $s_1$ extends $s$. So let $\reductionl
{t_2} {s_1} {v_2} {s_{12}} {j_2}$ with $v_2 \in \relv {k-j_2-1}
{\Gamma_{12}} {s_{12}} {T_2}$.

@@

@@

\subsubsection{Substitution Lemma}

The substitution lemma states that if (1) $v \in \relv {k_2}
{\Gamma_{12}} {s_{12}} {T_2}$ and (2) $t \in \rele {k_1} {\Gamma_1
  \envplus{{x} : S}} {s_1} T$ and (3) $\Gamma \ts {T_2} \sub S$ with
(4) $x \not\in \fn(T)$ and $\Gamma_1$ extends $\Gamma$ and
$\Gamma_{12}$ included in $\Gamma_1$ and $s_{12}$ extends $s_1$ and
$\Gamma_1$ agrees with $s_1$ and $\Gamma_{12}$ agrees with $s_{12}$
and a prefix of $\Gamma_{12}$ agrees with $s_1$, then $\subst {v} {x}
{t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}} {s_{12}} T$.

\begin{proofsketch}
By (1) and (3), it should hold that (5) $v \in \relv {k_2}
{\Gamma_{12}} {s_{12}} {S}$ by the subset semantics lemma.  Since (2)
holds, it should also hold that ${t} \in \rele {\min(k_1,k_2)}
{\Gamma_{12} \envplus{{x} : S}} {s_{12}} T$ by the extended
monotonicity lemma. Then, we can instantiate $x$ in the complete store
to map to what $v$ maps to. This should be fine by (5) and
monotonicity. Thus, ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12}
  \envplus{{x} : S}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}}
T$. Thanks to (4), we don't actually need $x$ to be held abstract in
the environment, because it won't occur in $T$ or its expansion, so we
can use the actual type constructor of $v$, say $T_{v}$, instead of
$S$ for $x$: ${t} \in \rele {\min(k_1,k_2)} {\Gamma_{12} \envplus{{x}
    : {T_{v_2}}}} {s_{12} \envplus{{x} \mapsto s_{12}(v)}} T$. This
implies what needs to be shown.

@@

\subsubsection{Subset Semantics Lemma}

The subset semantics lemma states that if (1) $v \in \relv k \Gamma s S$
and (2) $\Gamma \ts S <: U$, then $v \in \relv k \Gamma s U$.

\begin{proofsketch}
Because $S$ is a subtype of $U$, it should hold that the expansion of
$S$ subsumes the expansion of $U$. Therefore, for $v \in \relv k
\Gamma s U$, we have fewer and ``simpler'' definitions to check than
for $v \in \relv k \Gamma s S$. One worry is that for methods, because
the parameter type is contravariant, $U$ could require a more precise
type than $S$. This shouldn't lead to narrowing issues, because if the
parameter type becomes ill-typed, then the check is vacuously true.
\end{proofsketch}

\subsubsection{Extended Monotonicity Lemma}

The extended monotonicity lemma states that if $t \in \rele k {\Gamma
  \envplus{x : S}} s T$ then $t \in \rele j {\Gamma' \envplus{x : S}}
{s'} T$ for $j \leq k$, $\Gamma'$ extends $\Gamma$, $s'$
extends $s$, and $\Gamma$ agrees with $s$ and a prefix of $\Gamma'$
agrees with $s$.

\begin{proofsketch}
For the monotonicty with regards to the step-index, this follows
directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For the
environment and the store, this follows from the definition of $\rels
k {\Gamma \envplus{x : S}} s$. To extend the environment and the store
for $x : S$, we can append as much as we want to $\Gamma$ and $s$, to
get $\Gamma'$ and $s'$, and then ignore the last element which is for
$x : S$.
\end{proofsketch}
